\bf An observational study of papers published in \emph{PLOS ONE} and studies posted to a trial registry.

### 2021-04-23 manuscript/asa_template/asa_template.tex
revise manuscript anzctr plos

We applied the same text cleaning to both data sources. Text cleaning
aimed to standardise notation and statistical terminology, whilst
minimising changes to article style and formatting. \emph{R} code used
for data extraction and cleaning is available from
\url{https://github.com/agbarnett/stats_section}.

Mathematical notation including Greek letters was converted from Unicode
characters to plain text. For example, the Unicode characters
corresponding to \(\theta\) (\textless U+03B8\textgreater) were replaced
with `theta'. Similarly, common symbols outside of Unicode blocks
including `\%' (percent) and `\textless{}' (`less-than') were converted
into plain text using the `textclean' package \citep{textclean}. General
formatting was removed, this included carriage returns, punctuation
marks, in-text references (e.g.~``{[}42{]}'') centred equations, and
other non-ASCII characters. Text contained inside brackets was retained
to maximise content for analysis, with brackets removed.

We compiled an extensive list of statistical terms to standardise
descriptions of statistical methods reported across both datasets. An
initial list was compiled by calculating individual word frequencies and
identifying relevant terms that appeared at least 100 times. Further
terms were sourced from index searches of three statistics textbooks
\citep[\citet{Diggle2013},\citet{Bland2015}]{Dobson2018}. The final list
is provided as Supplementary Material. Plurals (e.g., `chi-squares')
unhyphenated (e.g., `chi square') and combined (e.g.~`chisquare') terms
were transformed to singular, hyphenated form (e.g., `chi-square').
Common statistical tests were also hyphenated (e.g., `hosmer lemeshow'
to `hosmer-lemeshow').

@@

We applied the clustering algorithm to the cleaned dataset, varying the
number of clusters from 1 to 50.

Results were transformed to lower case for the clustering, but examples
are given using the original capitalisation.

For each dataset, records assigned to invididual clusters were examined
for evidence of boilerplate text in two ways. We first reviewed the top
ten results that represented the strongest matches to each cluster.
Records assigned to the same cluster were also compared by calculating
pairwise cosine similarities; higher scores denoted a higher degree of
similarity in text between a pair of records.

@@

API searches returned 131,847 results (DOIs), of which 111,731 (85\%)
included a statistical methods section based on our search strategy. In
the final sample, 95,518 (85\%) DOIs returned an exact match against
common section headings: 64,133 for `statistical analysis', 13,380 for
`statistical analyses' and 13,627 for `data analysis'.

Among DOIs that did not meet the partial matching criteria, initial
search terms appeared in {[}TODO{]}.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The
total number of API search results peaked at volumes 8 (n = 19,045) and
9 (n = 19,045), corresponding to years 2013 and 2014. This trend aligned
with the total number of papers published in \emph{PLOS ONE} over the
same period. The percentage of records that included a statistical
methods section by volume based on our proposed matching criteria varied
between 64\% (volume 2) and 86\% (volume 9).

@@

{\centering \includegraphics[width=0.7\linewidth]{figures/plos.wordclouds}

@@

\caption{\label{fig:plos-clouds}Topic clouds for ten clusters for PLOS ONE (left) and ANZCTR (right)}\label{fig:unnamed-chunk-5}
\end{figure}

The topic clouds based on ten clusters are in
Figure\textasciitilde{}\ref{fig:plos-clouds}. Frequently occurring words
reflected the use of statistical software (Topics 3 and 5), descriptive
statistics (Topic 6), group based hypothesis testing (Topics 1 and 4)
and definitions of statistical significance (Topics\textasciitilde1 and
9). Statistical methods sections associated with regression (Topic 2)
and meta-analysis (Topic 7) were also identified.

\includegraphics[width=3.75in,height=2.75in,keepaspectratio]{asa_template_files/figure-latex/unnamed-chunk-6-1.png}

Topics related to the use of statistical software differentiated between
Prism GraphPad (Topic\textasciitilde3: n = 9,879; 8.8\%) and SPSS (Topic
5: n = 9,574; 8.6\%) . Manual review of results showed that nine out of
the ten top matching DOIs for Topic 3 stated the use of Prism GraphPad,
but did not specify which statistical methods were used. Top matching
sections for Topic 5 included information on SPSS version numbers and
definitions of statistical significance. Examples of boilerplate text
for selected topics are presented in Table
\ref{tab:plos-example-boilerplate}. Examples correspond to top ranking
DOIs, and DOIs with the higher number of matches based on cosine
similarity score greater than 0.8. Further results from cosine
similarity scores are provided in Supplementary File 2.

\includegraphics[width=8.00in,height=4.12in,keepaspectratio]{asa_template_files/figure-latex/unnamed-chunk-7-1.png}

Definitions of statistical significance featured strongly in Topic 1 (n
= 3,775) and Topic 9 (n = 6,104) combined with hypothesis testing for
comparing differences between two groups. Topic 1 reflected applications
of Student's t-test assuming a 5\% level of statistical significance.
Topic 9 referenced similar methods combined with multiple thresholds for
declaring statistical significance by asterisk: ``\(^{*}p<0.05\),
\(^{**}p<0.01\) and \(^{***}p<0.001\)''., a practice that has been
criticised {[}add wasserstein2019{]}.

Group-based hypothesis testing was a recurring theme across topics, with
text descriptions varying based on method(s) used. One-way analysis of
variance also featured strongly in Topic 4 (n = 10,163), with text
citing common methods for performing post-hoc multiple comparisons. Top
matching sections across these topics did not include definitions of
groups being compared.

Frequently occurring words in Topic 6 (n = 4,746) including `Mean', `SD'
and `SEM' reflected the use of descriptive statistics for summarising
continuous variables. Sections associated with this topic appeared to be
expanded versions of Topics 1 and 9, showing evidence of boilerplate
text in the form of descriptive statistics followed by hypothesis
testing using Student's t-test, Mann Whitney U or one-way analysis of
variance.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure\textasciitilde{}\ref{fig:anzctr-missing}. Of the 12,700
included studies, 9,523 (75\%) had a statistical methods section.

@@

\caption{ANZCTR search results.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{ANZCTR topic clouds.}\label{fig:unnamed-chunk-9}
\end{figure}

The median length of the section was 129 words with an inter-quartile
range of 71 to 219 words. Some methods sections were only one word,
including ``ANOVA'', ``t-test'', ``SPSS'' and even ``SSPS''.

The clustering algorithm found groups that were purely sample size
calculations (topic 2), pilot studies (topic 5), safety/tolerability
studies (topic 6) and repeated measures ANOVA (topic 10). There were
cases where the exact same method section had been re-used in a
different study.

We also found evidence of `boilerplate' sections clustered as topic 3,
example text

\textbf{Box x: Examples of boilerplate text from ANZCTR}

Topic 3

\begin{itemize}
\tightlist
\item
  ``Shapiro Wilk test was used as normality test. Continuous variables
  were compared using Student t-test and Mann-Whitney U test when the
  data were not normally distributed. Categorical variables were
  compared using Pearson's chi-squared test and Fisher's exact test.
  Paired data were analyzed using Paired t-test and Wilcoxon signed rank
  test when data were not normally distributed.''
\item
  ``Comparisons between categorical variables will be made either using
  chi square or Fisher exact test. Continuous data will be compared
  using the Student's t-test or Mann-Whitney U test. Two sided p values
  of less than 0.05 will be considered statistically significant.''
\end{itemize}

We examined if four study characteristics were associated with a missing
statistics section. The odds ratios and 95\% credible intervals are in
Table\textasciitilde X. Observational studies were less likely to have a
missing methods section compared with interventional studies. Missing
sections became less likely over time. Studies with more funders and a
larger target sample size were less likely to have a missing methods
section.

\includegraphics[width=4.66in,height=1.55in,keepaspectratio]{asa_template_files/figure-latex/unnamed-chunk-10-1.png}

\begin{itemize}
\tightlist
\item
  Final sample size
\end{itemize}

@@

\section{Supplementary}

\subsection{PLOS}

{\centering \includegraphics[width=1\linewidth]{figures/supplementary/supp_plosone.50clusters}

\caption{PLOS ONE clustering metrics.}\label{fig:unnamed-chunk-11}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/supplementary/plos.topicvalues}

\caption{PLOS ONE topic values.}\label{fig:unnamed-chunk-12}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/supplementary/supp_anzctr.50clusters}

\caption{ANZCTR clustering metrics.}\label{fig:unnamed-chunk-13}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/supplementary/anzctr.topicvalues}

\caption{ANZCTR topic values.}\label{fig:unnamed-chunk-14}
\end{figure}

>>>>

We applied the same text cleaning steps to both data sources using
functions from the `textclean package' \citep{textclean}. Text cleaning
aimed to standardise notation and statistical terminology, whilst
minimising changes to article style and formatting. \emph{R} code used
for data extraction and cleaning is available from
\url{https://github.com/agbarnett/stats_section}.

Mathematical notation including Greek letters was converted from Unicode
characters to plain text. For example, the Unicode character
corresponding to \(\theta\) (\textless U+03B8\textgreater) was replaced
with `theta'. Common symbols outside of Unicode blocks including `\%'
(percent) and `\textless{}' (`less-than') were similarly converted into
plain text. General formatting was removed, this included carriage
returns, punctuation marks, in-text references (e.g.~``{[}42{]}'')
centred equations, and other non-ASCII characters. Text contained inside
brackets was retained to maximise content for analysis, with brackets
removed.

We compiled an extensive list of statistical terms to standardise
descriptions of statistical methods reported across both datasets. An
initial list was compiled by calculating individual word frequencies and
identifying relevant terms that appeared at least 100 times. Further
terms were sourced from index searches of three statistics textbooks
\citep[\citet{Diggle2013},\citet{Bland2015}]{Dobson2018}. The final list
is provided in Supplementary File 1. Plurals (e.g., `chi-squares')
unhyphenated (e.g., `chi square') and combined (e.g.~`chisquare') terms
were transformed to singular, hyphenated form (e.g., `chi-square').
Common statistical tests were also hyphenated (e.g., `hosmer lemeshow'
to `hosmer-lemeshow').

@@

We applied the clustering algorithm to the cleaned dataset, varying the
number of clusters from 1 to 50. Clustering solutions were assessed
using silhouette score, within-cluster dispersion and between-cluster
dispersion.

For each dataset, records assigned to invididual clusters were examined
for evidence of boilerplate text in two ways. We first reviewed the
strongest matches to each cluster based on topic scores{[}may not be the
right word{]}. We also compared similarities in text between pairs of
records assigned to each cluster. Similarities were determined from
calculating cosine similarity scores, with comparisons limited to the
top 500 records per cluster. Higher similarity score represent greater
overlap in text content between statistical methods sections. Results
were transformed to lower case for the clustering, but examples are
given using the original capitalisation.

@@

API searches returned 131,847 results (DOIs) based on chosen search
terms. After partial matching, 111,731 (85\%) statistical methods
sections were identified. In the final sample, 95,518 (85\%) DOIs
returned an exact match against common section headings: 64,133 for
`statistical analysis', 13,380 for `statistical analyses' and 13,627 for
`data analysis'. For our random sample of DOIs that did not meet the
partial matching criteria, initial search terms appeared in Introduction
(number), Results (number) only. xxx appeared in the Methods section
without a section heading and xxx had a non-standard section heading.
{[}TODO{]}.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The
total number of API search results peaked at volumes 8 (n = 19,045) and
9 (n = 19,045) in years 2013 and 2014. This trend aligned with the total
number of papers published in \emph{PLOS ONE} over the same period. The
percentage of records that included a statistical methods section by
volume based on our proposed matching criteria varied between 64\%
(volume 2) and 86\% (volume 9).

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/plos.wordclouds}

@@

\caption{\label{fig:plos-clouds}Topic clouds for ten clusters for PLOS ONE}\label{fig:unnamed-chunk-5}
\end{figure}

The topic clouds based on ten clusters are in Figure
\ref{fig:plos-clouds}. Frequently occurring words reflected the use of
statistical software (topics 3 and 5), descriptive statistics (topic 6),
group based hypothesis testing (topics 1 and 4) and definitions of
statistical significance (topics 1 and 9). Statistical methods sections
related to regression (topic 2) and meta-analysis (topic 7) were also
identified.

Topics related to statistical software differentiated between Prism
GraphPad (topic 3: n = 9974; 8.6\%) and SPSS (topic 5: n = 9648; 8.3\%)
. Manual review of results showed that nine out of the ten top matching
DOIs for topic 3 stated the use of Prism GraphPad, but did not specify
which statistical methods were used. Top matching sections for topic 5
included information on SPSS version numbers and definitions of
statistical significance.

\begin{landscape}
\begin{table}[]
\centering
\caption{Example boilerplate text from PLOS ONE dataset}
\label{tab:plos-example-boilerplate}
\begin{tabular}{p{0.1\linewidth} p{0.9\linewidth}}
\hline
\textbf{Topic} & \textbf{Statistical methods section} \\ \hline
1 & Student’s t-test was used for statistical analysis. A p value of $<$0.05 was considered statistically significant. \\
& Statistical analysis was performed using the Students t-test. P values of less than 0.05 were considered significant.\\
\hline
3 & GraphPad Prism (Graphpad Software, San Diego, CA) was used for all analyses.\\
 & All statistical analysis was performed using Graphpad Prism software.\\
 \hline
4 & Significant differences were determined using analysis of variance (ANOVA) followed by Tukey post-hoc tests for multiple comparisons.\\
& Data are expressed as the mean ± SEM. Statistical analysis was performed using one-way analysis of variance (ANOVA) test, followed by Dunnett’s multiple comparison post hoc test. P values less than 0.05 were considered significant.\\
\hline
5 & SPSS software version 17.0 (SPSS, Chicago, IL, USA) was used for statistical analysis.\\
  & The results are presented as the mean ± SEM. Data were analyzed by one-way ANOVA and LSD tests using the SPSS 13.0 software (SPSS Inc., Chicago, IL, USA). The difference was considered statistically significant at P$<$0.05.\\
  \hline
6 & All results are expressed as means ± standard deviation (± SD).\\
& Data are expressed as mean ± Standard error of the mean (SEM). Statistical analysis of the data was performed by the Student’s t-test. A value of P$<$0.05 was considered statistically significant.\\
\hline
9 & Statistical significance was determined by Students t-tests. $^{*} P<0.05$, $^{**} P = <0.01$, $^{***} P<0.001$.\\
 & Data are presented as the mean ± SD. Statistical analysis was performed using the Students unpaired t-test. Differences were considered statistically significant at $^{*}p<0.05$, $^{**}p<0.01$ and $^{***}p<0.001$.\\
 \hline
\end{tabular}
\end{table}
\end{landscape}

Definitions of statistical significance featured strongly in topic 1 (n
= 3784; 3.2\%) and topic 9 (n = 6195; 5.3\%), combined with text
describing hypothesis testing for comparing differences between two
groups. Topic 1 reflected applications of Student's t-test assuming a
5\% level of statistical significance. Topic 9 referenced similar
methods combined with multiple thresholds for declaring statistical
significance by asterisk: ``\(^{*}p<0.05\), \(^{**}p<0.01\) and
\(^{***}p<0.001\)'', a practice that has been criticised
\citep{Wasserstein2019}.

Group-based hypothesis testing was a recurring theme across topics, with
text descriptions varying based on method(s) used. One-way analysis of
variance also featured strongly in topic 4 (n = 10212; 8.8\%), combined
with common methods for performing post-hoc multiple comparisons.
Frequently occurring words in topic 6 (n = 4764; 4.1\%) reflected
mentions of descriptive statistics for summarising continuous variables,
for example: ``Data are expressed as means ± standard error of the mean
(SEM).'' Sections assigned to this topic appeared to be expanded
versions of topics 1 and 9. Examples of boilerplate text were in the
form of descriptive statistics followed by hypothesis testing, for
example, Student's t-test, Mann Whitney U or one-way analysis of
variance.

Examples of boilerplate text for selected topics are presented in Table
\ref{tab:plos-example-boilerplate}. Detailed findings based on cosine
similarity scores are provided in Supplementary File 2.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure \ref{fig:anzctr-missing}. Of the 12,700 included studies,
9,523 (75\%) had a statistical methods section. The median length of
sections was 129 words with an inter-quartile range of 71 to 219 words.

@@

\caption{\label{fig:anzctr-missing}ANZCTR search results.}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{ANZCTR topic clouds.}\label{fig:unnamed-chunk-8}
\end{figure}

The clustering algorithm found groups that were purely sample size
calculations (topic 2, n = 1834; topic 4, n = 909), pilot studies (topic
5, n = 834), safety/tolerability studies (topic 6, n = 524),
intervention studies (topic 8, n = 1020) and repeated measures ANOVA
(topic 10 = 852). Examples of boilerplate text are provided in Table
\ref{tab:anzctr-example-boilerplate} and Supplementary File 3.

Some methods sections were only one word, including ``ANOVA'',
``t-test'', ``SPSS'' and even ``SSPS''. There were cases where the exact
same method section had been re-used in a different study. For example,
in topic 7 (n = 333), 232 sections stated `descriptive statistics' or
`descriptive statistics used' with no additional details provided. Other
instances with expanded descriptions of methods included topic 3 (n =
1277) outlining descriptive analyses only, and topic 6 (n = 524), for
hypothesis testing methods, statistical significance and software.

In other cases, text had been slightly modified to account for changes
in primary and secondary outcomes. Examples of these text changes were
found in topic 2 (n = 1834) and topic 4 (n = 909); identified instances
related to sample size calculations for patient recruitment to different
studies (see Supplementary File 3).

\begin{landscape}
\begin{table}[]
\centering
\caption{Example boilerplate text from ANZCTR dataset}
\label{tab:anzctr-example-boilerplate}
\begin{tabular}{p{0.1\linewidth} p{0.9\linewidth}}
\hline
\textbf{Topic} & \textbf{Statistical methods section} \\ \hline
3 & Comparisons between categorical variables will be made either using chi square or Fisher exact test. Continuous data will be compared using the Student’s t-test or Mann-Whitney U test. Two sided p values of less than 0.05 will be considered statistically significant.\\
 &  The Mann-Whitney U, Student t, 1-way ANOVA, and Kruskal-Wallis tests will be used to compare continuous variables where relevant. The Fisher exact and Pearson’s Chi-square test will be used to compare proportions as appropriate. \\ \hline
5 & Pilot study \\
 &  No formal sample size calculation was performed \\ \hline
7 & Descriptive statistics \\
 &  Descriptive statistics used \\ \hline
9 & Linear mixed models will be used to analyse the data. \\ \hline
10 & Repeated measures of ANOVA \\
 &  Pre-, during, post- and follow-up variables will be subjected to mixed methods and repeated measures analyses to determine significant changes over (group and) time. \\ \hline
\end{tabular}
\end{table}
\end{landscape}

We examined if four study characteristics were associated with a missing
statistics section. The odds ratios and 95\% credible intervals are in
Table \ref{tab:anzctr-missing-odds}. Observational studies were less
likely to have a missing methods section compared with interventional
studies. Missing sections became less likely over time. Studies with
more funders and a larger target sample size were less likely to have a
missing methods section.

\begin{table}[]
\centering
\caption{Regression results for factors associated with missing statistical methods sections in ANZCTR}
\label{tab:anzctr-missing-odds}
\begin{tabular}{lcc}
\hline
Variable & Odds ratio & 95$\%$ CI \\
\hline
Study type $=$ Observational & 0.78 & (0.69, 0.89) \\
Date (per year) & 0.90 & (0.88, 0.91) \\
Number of funders & 0.80 & (0.74, 0.86) \\
Target sample size (per doubling) & 0.90 & (0.88, 0.92)\\
\hline
\end{tabular}
\end{table}

@@
